Recursive Solutions to First-Order Model Counting

### 2022-08-02 doc/paper/paper.tex
Minor changes

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic (FOL) given the size(s) of its
domain(s) \citep{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as a WMC instance---in fact, the WFOMC algorithm
by \citet{DBLP:conf/ijcai/BroeckTMDR11} can do so when it is unable to solve the
instance in a lifted manner. WFOMC emerged as the dominant approach to
\emph{lifted (probabilistic) inference}. Lifted inference techniques exploit
symmetries in probabilistic models by reasoning about sets rather than
individuals \citep{DBLP:conf/ecai/Kersting12}. By doing so, many instances
become solvable in polynomial time \citep{DBLP:conf/nips/Broeck11}. Lifted
inference algorithms are typically used on probabilistic models such as
probabilistic programming languages
\citep{DBLP:journals/ml/RaedtK15,DBLP:journals/ijar/RiguzziBZCL17}, Markov logic
networks
\citep{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16,DBLP:journals/ml/RichardsonD06},
and other lifted graphical \citep{DBLP:journals/ml/KimmigMG15} and statistical
relational \citep{DBLP:series/synthesis/2016Raedt} models. Lifted inference
techniques for probabilistic databases, while developed somewhat independently,
have also been inspired by WFOMC
\citep{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.

@@

Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the size of the instance. The
equivalent notion in (W)FOMC is liftability. A (W)FOMC instance is
\emph{(domain-)liftable} if it can be solved in time polynomial in the size(s)
of the domain(s) \citep{jaeger2012liftability}. Over more than a decade, more
and more classes of instances were shown to be liftable
\citep{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, \citet{DBLP:conf/nips/Broeck11} showed that the class of all sentences of
FOL with up to two variables (denoted \FOtwo{}) is liftable. Then
\citet{DBLP:conf/pods/BeameBGS15} proved that there exists a sentence with three
variables for which FOMC is $\#\P_{1}$-complete (i.e., \FOthree{} is not
liftable). Since these two results came out, most of the research on (W)FOMC
focused on developing faster solutions for the \FOtwo{} fragment
\citep{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining
new liftable fragments. These fragments include \SFO{} and \SRU{}
\citep{DBLP:conf/nips/KazemiKBP16}, \Uone{} \citep{DBLP:conf/lics/KuusistoL18},
\FOtwo{} with tree axioms \citep{DBLP:conf/kr/BremenK21}, and \Ctwo{} (i.e., the
two variable fragment with counting quantifiers)
\citep{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. On the
empirical front, there are several open-source implementations of exact WFOMC
algorithms: \textsc{ForcLift} \citep{DBLP:conf/ijcai/BroeckTMDR11},
probabilistic theorem proving \citep{DBLP:journals/cacm/GogateD16}, and
\textsc{L2C} \citep{DBLP:conf/kr/KazemiP16}. Approximate counting is supported
by \textsc{ApproxWFOMC} \citep{DBLP:conf/ijcai/BremenK20} as well as
\textsc{ForcLift} \citep{DBLP:conf/uai/BroeckCD12} and probabilistic theorem
proving \citep{DBLP:journals/cacm/GogateD16}.

@@

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances (e.g., counting
various injective mappings) in a lifted manner. The ideas presented in this
paper are implemented in \textsc{Crane}---an extension of the arguably most
well-known WFOMC algorithm \textsc{ForcLift} written in Scala. \textsc{ForcLift}
works in two stages: compilation and evaluation/propagation.\footnote{There is
  also an intermediate stage called \emph{smoothing} that takes place between
  compilation and evaluation. As our changes to smoothing are rather elementary,
  we do not discuss them to not distract the reader from the main contributions
  of this paper.} In the first part, various \emph{(compilation) rules} are
applied to the input (or some derivative) formula, gradually constructing a
circuit. In the second part, the weights of the instance (and sometimes
constants) are propagated through the circuit, computing the WMC\@.

@@

The main conceptual difference between \textsc{Crane} and \textsc{ForcLift} is
that we utilise labelled directed graphs instead of circuits. The cycles in
these graphs represent recursive calls. Suppose the input formula $\phi$ depends
on a domain of size $n \in \mathbb{N}$. \emph{Generalised domain recursion}
(GDR), one of the new compilation rules, transforms $\phi$ into a different
formula $\psi$ that has an additional constant and some new \emph{constraints}.
After some additional transformations, the constraints in $\psi$ become
`uniform' and can be removed, replacing the domain of size $n$ with a new domain
of size $n-1$---this is the responsibility of the \emph{constraint removal} (CR)
compilation rule. Afterwards, another new compilation rule recognizes that the
resulting formula $\phi(n-1)$ is just like the input formula $\phi(n)$ except it
refers to a different domain. This observation allows us to add a cycle-forming
edge to the graph, which can be interpreted as function $f$ relying on $f(n-1)$
to compute $f(n)$.

@@

Our representation of FOMC instances is largely based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by
\citet{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} can translate sentences
in a variant of function-free many-sorted FOL with equality to this internal
format. An \emph{atom} is $p(t_1, \dots, t_n)$ for some predicate $p$ and terms
$t_{1}, \dots, t_{n}$. Here, $n \in \mathbb{N}_0$ is the \emph{arity} of $p$. A
\emph{term} is either a constant or a variable. A \emph{literal} is either an
atom or the negation of an atom (denoted by $\neg p(t_1, \dots, t_n)$). Let
$\mathcal{D}$ be the set of all domains. Initially, $\mathcal{D}$ contains all
domains mentioned by the input (W)FOMC instance. During compilation, new domains
are added to $\mathcal{D}$ by some of the compilation rules.

@@

Once domains are mapped to finite sets and constants to specific (and different)
elements in those sets, a formula can be viewed as a set of conditions that the
predicates (interpreted as relations) have to satisfy.\footnote{If some domain
  is not big enough to contain all of its constants, the formula is
  unsatisfiable.} Hence, FOMC is the problem of counting the number of
combinations of relations that satisfy these conditions.

@@

\paragraph*{Notation.}
We write $\langle\rangle$ and $\langle x \rangle$ to denote an empty list and a
list with one element $x$, respectively, and $|l|$ for the length of list $l$.
We write $\pfun$ for partial functions. Let $S$ be a set of constraints or
literals, $V$ a set of variables, and $x$ a variable or a constant. We write
$S[x/V]$ to denote $S$ with all occurrences of all variables in $V$ replaced
with $x$.\footnote{Note that if $(X, Y)$ is a two-variable constraint,
  substituting a constant $c$ for $X$ would result in $(c, Y)$, which would have
  to be rewritten as $(Y, c)$ to fit the definition of a constraint.}

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[every node/.style={draw,ellipse},edge from
    parent/.style={draw,-Latex},sibling distance=3cm]
    \node (dr) {$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$}
    child {node {$\bigvee_{\Delta^\top \subseteq \Delta}$}
      child {node {$\CR_{\Gamma \mapsto \Gamma'}$}
        child {
          node[circle] {$\wedge$}
          child {node {$\bot$}}
          child {node (ref) {$\Reff_{\{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^{\bot} \,\}}$}}
        }}};
    \draw[-Latex, bend right] (ref) to (dr);
  \end{tikzpicture}
  \caption{A simplified version of an FCG constructed by \textsc{Crane} for the
    problem of counting partial injections from \cref{example:first}. Label
    $\bigvee_{\Delta^\top \subseteq \Delta}$ denotes set-disjunction, $\land$
    denotes conjunction, and $\bot$ denotes a contradiction---see the work by
    \citet{DBLP:conf/ijcai/BroeckTMDR11} for the descriptions of these node
    types. Here we omit nodes whose only arithmetic effect is multiplication by
    one. Some of these nodes play an important role in the weighted version of
    the problem, whereas others are remnants of the interaction between
    compilation rules and the way in which \textsc{ForcLift} handles existential
    quantifiers.}\label{fig:examplefcg}
\end{figure}

% From circuits to graphs.
A \emph{first-order deterministic decomposable negation normal form
  computational graph} (FCG) is a (weakly connected) directed graph with a
single source, node labels, and ordered outgoing edges.\footnote{Note that
  imposing an ordering on outgoing edges is just a limited version of edge
  labelling.} Node labels consist of two parts: the \emph{type} and the
\emph{parameters}. The type of a node determines its out-degree. Similarly to
\citet{DBLP:conf/ijcai/BroeckTMDR11}, we write $T_p$ for an FCG that has a node
with the label $T_p$ (i.e., type $T$ and parameter(s) $p$) and $\star$'s as all
of its direct successors.

Most node types are as in previous work
\citep{DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11}. The type for
non-tree-like edges (denoted by $\Reff$), while used before, is extended to
contain the information necessary to support recursive calls. We also add three
new types:
\begin{itemize}
  \item a type for constraint removal denoted by $\CR$,
  \item a type for generalised domain recursion denoted by $\GDR$ (both with
        out-degree one),
  \item and $\star$---a placeholder type (with out-degree zero) for nodes that
        are going to be replaced.
\end{itemize}
When drawing an FCG, we order outgoing edges from left to right, write node
labels directly on the nodes, and omit irrelevant labels and/or parameters. See
\cref{fig:examplefcg} for an example FCG\@. Its source node has out-degree 1,
label $\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$, and type $\GDR$.

% Connecting FCGs with formulas
Finally, we introduce a structure that represents a solution to a (W)FOMC
problem while it is still being built. A \emph{chip} is a pair $(G, L)$, where
$G$ is an FCG, and $L$ is a list of formulas, s.t. $|L|$ is equal to the number
of $\star$'s in $G$. The intuition behind this definition is that $L$ contains
formulas that still need to be compiled. Once a formula is compiled, it replaces
one of the $\star$'s in $G$ according to a set order. We say that an FCG is
\emph{complete} (i.e., it represents a \emph{complete solution}) if it has no
$\star$'s. Similarly, a chip is complete if its FCG is complete (or,
equivalently, the list of formulas is empty).

@@

A \emph{(compilation) rule} takes a formula and returns a set of chips. The
cardinality of this set is the number of different ways in which the rule can be
applied to the input formula. While \textsc{ForcLift}
\citep{DBLP:conf/ijcai/BroeckTMDR11} heuristically chooses one of them, in an
attempt to not miss a solution, \textsc{Crane} returns them all. In particular,
if a rule returns an empty set, then that rule does not apply to the formula,
and the algorithm continues with the next rule.

@@

% Description of GDR, in contrast to DR
In contrast, GDR has only one precondition: for GDR to be applicable on domain
$\Omega \in \mathcal{D}$, there must be at least one variable with domain
$\Omega$ that is featured in a literal (and not just in constraints). Without
such variables, GDR would have no effect on the formula. GDR is also simpler in
that the expanded formula is left as-is to be handled by other compilation
rules. Typically, after a few more rules are applied, a combination of $\CR$ and
$\Reff$ nodes introduces a loop back to the $\GDR$ node, thus completing the
definition of a recursive function. The GDR compilation rule is summarised as
\cref{alg:domainrecursion} and explained in more detail using the example below.

@@

\begin{example}
  Let $\phi = \{\, c_1, c_2, c_3 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &= (\emptyset, \{\, (Y, X) \,\}, \{\, X \mapsto \Delta^\top, Y \mapsto \Delta^\top \,\}), \\
    c_2 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (X, x), (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
          \end{multlined}\\
    c_3 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \\
            \{\, (X, x), (Z, X), (Z, x) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Gamma \,\}).
          \end{multlined}
  \end{align*}
  Domain $\Gamma$ and its element $x \in \Gamma$ satisfy the preconditions for
  CR\@. The rule introduces a new domain $\Gamma'$ and transforms $\phi$ to
  $\phi' = (c_1', c_2', c_3')$, where
  \begin{align*}
    c_1' &= c_1, \\
    c_2' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
           \end{multlined} \\
    c_3' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (Z, X) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Gamma' \,\}).
           \end{multlined}
  \end{align*}
\end{example}

@@

Let $\Doms$ be a function that maps any clause or formula to the set of domains
used within, and let $\dom(f)$ denote the domain of function $f$.

We use (integer-valued) hash functions to efficiently discard pairs of formulas
that are too different for recursion to be established. The hash code of a
clause $c = (L, C, \delta_{c})$ (denoted by $\# c$) combines the hash codes of
the sets of constants and predicates in $c$, the numbers of positive and
negative literals, the number of inequality constraints $|C|$, and the number of
variables $|\Vars(c)|$. The hash code of a formula $\phi$ combines the hash
codes of all its clauses and is denoted by $\#\phi$.

\citet{DBLP:conf/ijcai/BroeckTMDR11} use a cache to check if a formula is
identical to one of the formulas that have already been fully compiled. To
facilitate recursion, we extend the caching scheme to include formulas that have
been encountered but not fully compiled yet. Formally, we define a \emph{cache}
to be a map from integers (e.g., hash codes) to sets of pairs of the form
$(\phi, v)$, where $\phi$ is a formula, and $v$ is an FCG node.

@@

When \textsc{ForcLift} \citep{DBLP:conf/ijcai/BroeckTMDR11} compiles a WFOMC
instance into a circuit, each gate type encodes an arithmetic operation on its
inputs and parameters. These operations are then immediately performed while
traversing the circuit and using domain sizes and weights as the initial inputs.
With \textsc{Crane}, the interpretation of an FCG is a collection of functions.
Each function has (some) domain sizes as parameters and may contain recursive
calls to other functions, including itself. While there may be any number of
subsidiary functions, there is always one main function that can be called with
the sizes of the domains of the input formula as arguments. Henceforth, this
function is always called $f$, and it is defined by the source node.

@@

\begin{example}\label{example:interpretation}
  Let us use the FCG from \cref{fig:examplefcg} as an example. The input formula
  (i.e., the formula in \cref{example:first}) has two domains: $\Gamma$ and
  $\Delta$. Thus, the interpretation of the FCG is a function
  $f\colon \mathbb{N}_{0} \times \mathbb{N}_{0} \to \mathbb{R}_{\ge 0}$. Let
  $m \coloneqq |\Gamma|$, and $n \coloneqq |\Delta|$. The node labelled
  $\bigvee_{\Delta^{\top} \subseteq \Delta}$ tells us that
  $f(m, n) = \sum_{l = 0}^{n} \binom{n}{l} \square$, where $\square$ is the
  interpretation of the remaining subgraph, and $l$ iterates over all possible
  sizes of $\Delta^{\top}$. It also creates two subdomains
  $\Delta^{\top}, \Delta^{\bot} \subseteq b$ that partition $\Delta$, i.e., as
  the size of $\Delta^{\top}$ increases, the size of $\Delta^{\bot}$
  correspondingly decreases. Nodes labelled $\land$ correspond to
  multiplication. Therefore,
  $f(m, n) = \sum_{l = 0}^{n} \binom{n}{l} \diamondsuit \times \heartsuit$,
  where $\diamondsuit$ is the interpretation of the contradiction (i.e., $\bot$)
  node, and $\heartsuit$ is the interpretation of the $\Reff$ node.

@@

In this section, we compare \textsc{Crane} and \textsc{ForcLift}
\citep{DBLP:conf/ijcai/BroeckTMDR11} on their ability to count various kinds of
functions. (Other WFOMC algorithms such as \textsc{L2C}
\citep{DBLP:conf/kr/KazemiP16} and probabilistic theorem proving
\citep{DBLP:journals/cacm/GogateD16} are unable to solve any of the instances
that \textsc{ForcLift} fails on.) We begin by describing how such functions can
be expressed in FOL\@. \textsc{ForcLift} then translates these sentences in FOL
to formulas as defined in \cref{def:formula}.

>>>>

% in fact, the WFOMC algorithm by \citet{DBLP:conf/ijcai/BroeckTMDR11} can do so when it is unable to solve the instance in a lifted manner.

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic (FOL) given the size(s) of its
domain(s) \citep{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as an instance of (propositional) weighted model
counting \citep{DBLP:journals/ai/ChaviraD08}. WFOMC emerged as the dominant
approach to \emph{lifted (probabilistic) inference}. Lifted inference techniques
exploit symmetries in probabilistic models by reasoning about sets rather than
individuals \citep{DBLP:conf/ecai/Kersting12}. By doing so, many instances
become solvable in polynomial time \citep{DBLP:conf/nips/Broeck11}. Lifted
inference algorithms are typically used on probabilistic models such as
probabilistic programming languages
\citep{DBLP:journals/ml/RaedtK15,DBLP:journals/ijar/RiguzziBZCL17}, Markov logic
networks
\citep{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16,DBLP:journals/ml/RichardsonD06},
and other lifted graphical \citep{DBLP:journals/ml/KimmigMG15} and statistical
relational \citep{DBLP:series/synthesis/2016Raedt} models. Lifted inference
techniques for probabilistic databases, while developed somewhat independently,
have also been inspired by WFOMC
\citep{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.

@@

Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the size of the instance. The
equivalent notion in (W)FOMC is liftability. A (W)FOMC instance is
\emph{(domain-)liftable} if it can be solved in time polynomial in the size(s)
of the domain(s) \citep{jaeger2012liftability}. Over more than a decade, many
classes of instances were shown to be liftable
\citep{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, \citet{DBLP:conf/nips/Broeck11} showed that the class of all sentences of
FOL with up to two variables (denoted \FOtwo{}) is liftable. Then
\citet{DBLP:conf/pods/BeameBGS15} proved that there exists a sentence with three
variables for which FOMC is $\#\P_{1}$-complete (i.e., \FOthree{} is not
liftable). Since these two results came out, most of the research on (W)FOMC
focused on developing faster solutions for the \FOtwo{} fragment
\citep{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining new
liftable fragments. These fragments include \SFO{} and \SRU{}
\citep{DBLP:conf/nips/KazemiKBP16}, \Uone{} \citep{DBLP:conf/lics/KuusistoL18},
\FOtwo{} with tree axioms \citep{DBLP:conf/kr/BremenK21}, and \Ctwo{} (i.e., the
two-variable fragment with counting quantifiers)
\citep{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. On the
empirical front, there are several open-source implementations of exact WFOMC
algorithms: \textsc{ForcLift} \citep{DBLP:conf/ijcai/BroeckTMDR11},
probabilistic theorem proving \citep{DBLP:journals/cacm/GogateD16}, and
\textsc{L2C} \citep{DBLP:conf/kr/KazemiP16}. Approximate counting is supported
by \textsc{ApproxWFOMC} \citep{DBLP:conf/ijcai/BremenK20} as well as
\textsc{ForcLift} \citep{DBLP:conf/uai/BroeckCD12} and probabilistic theorem
proving \citep{DBLP:journals/cacm/GogateD16}.

@@

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances (e.g., counting
various injective mappings) in a lifted manner. The ideas presented in this
paper are implemented in \textsc{Crane}---an extension of the arguably most
well-known WFOMC algorithm \textsc{ForcLift}. \textsc{ForcLift} works in two
stages: compilation and evaluation/propagation.\footnote{There is also an
  intermediate stage called \emph{smoothing} that takes place between
  compilation and evaluation. As our changes to smoothing are rather elementary,
  we do not discuss them to not distract the reader from the main contributions
  of this paper.} In the first stage, various \emph{(compilation) rules} are
applied to the input (or some derivative) formula, gradually constructing a
circuit. In the second stage, the weights of the instance are propagated through
the circuit, computing the weighted model count. Along with new compilation
rules, \textsc{Crane} introduces changes to both stages of the process. First,
while \textsc{ForcLift} applies compilation rules via greedy\footnote{The
  algorithm is not described in any paper on \textsc{ForcLift} but can be found
  in its source code at \url{https://dtai.cs.kuleuven.be/drupal/wfomc}} search,
\textsc{Crane} uses a hybrid search algorithm that applies some rules greedily
and some using breadth-first search. Second, the product of compilation is not
directly evaluated but rather interpreted as a collection of functions on domain
sizes.

@@

The main conceptual difference between \textsc{Crane} and \textsc{ForcLift} is
that we utilise labelled directed graphs instead of circuits, where cycles
represent recursive calls. Suppose the input formula $\phi$ depends on a domain
of size $n \in \mathbb{N}$. \emph{Generalised domain recursion} (GDR)---one of
the new compilation rules---transforms $\phi$ into a different formula $\psi$
that has an additional constant and some new \emph{constraints}. After some
additional transformations, the constraints in $\psi$ become `uniform' and can
be removed, replacing the domain of size $n$ with a new domain of size
$n-1$---this is the responsibility of the \emph{constraint removal} (CR)
compilation rule. Afterwards, another compilation rule recognizes that the
resulting formula is just like the input formula $\phi$ except it refers to a
different domain. This observation allows us to add a cycle-forming edge to the
graph, which can be interpreted as function $f$ relying on $f(n-1)$ to compute
$f(n)$.

% search
% interpretation

@@

Our representation of FOMC instances is largely based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by
\citet{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} can translate sentences
in a variant of function-free many-sorted FOL with equality to this internal
format. An \emph{atom} is $p(t_1, \dots, t_n)$ for some predicate $p$ and terms
$t_{1}, \dots, t_{n}$. A \emph{term} is either a constant or a variable. A
\emph{literal} is either an atom or the negation of an atom (denoted by
$\neg p(t_1, \dots, t_n)$). Let $\mathcal{D}$ be the set of all domains.
Initially, $\mathcal{D}$ contains all domains mentioned by the input (W)FOMC
instance. During compilation, new domains are added to $\mathcal{D}$ by some of
the compilation rules.

@@

Once domains are mapped to finite sets and constants to specific (and different)
elements in those sets, a formula can be viewed as a set of conditions that the
predicates (interpreted as relations) have to satisfy. Hence, FOMC is the
problem of counting the number of combinations of relations that satisfy these
conditions.

% \footnote{If some domain is not big enough to contain all of its constants,
% the formula is unsatisfiable.}

@@

\paragraph*{Notation.}
We write $\langle\rangle$ and $\langle x \rangle$ to denote an empty list and a
list with one element $x$, respectively, and $|l|$ for the length of list $l$.
We write $\pfun$ for partial functions. Let $S$ be a set of constraints or
literals, $V$ a set of variables, and $x$ a variable or a constant. We write
$S[x/V]$ to denote $S$ with all occurrences of all variables in $V$ replaced
with $x$.\footnote{If $(X, Y)$ is a two-variable constraint, substituting a
  constant $c$ for $X$ would result in $(c, Y)$, which would have to be
  rewritten as $(Y, c)$ to fit the definition of a constraint.}

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[every node/.style={draw,ellipse},edge from
    parent/.style={draw,-Latex},sibling distance=3cm]
    \node (dr) {$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$}
    child {node {$\bigvee_{\Delta^\top \subseteq \Delta}$}
      child {node {$\CR_{\Gamma \mapsto \Gamma'}$}
        child {
          node[circle] {$\wedge$}
          child {node {$\bot$}}
          child {node (ref) {$\Reff_{\{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^{\bot} \,\}}$}}
        }}};
    \draw[-Latex, bend right] (ref) to (dr);
  \end{tikzpicture}
  \caption{A simplified version of an FCG constructed by \textsc{Crane} for the
    problem of counting partial injections from \cref{example:first}. Label
    $\bigvee_{\Delta^\top \subseteq \Delta}$ denotes set-disjunction, $\land$
    denotes conjunction, and $\bot$ denotes a contradiction---see the work by
    \citet{DBLP:conf/ijcai/BroeckTMDR11} for the descriptions of these node
    types. Here we omit some parameters as well as nodes whose only arithmetic
    effect is multiplication by one. Some of these nodes play an important role
    in the weighted version of the problem, whereas others are remnants of the
    interaction between compilation rules and the way in which \textsc{ForcLift}
    handles existential quantifiers.}\label{fig:examplefcg}
\end{figure}

% From circuits to graphs.
A \emph{first-order deterministic decomposable negation normal form
  computational graph} (FCG) is a (weakly connected) directed graph with a
single source, node labels, and ordered outgoing edges.\footnote{Imposing an
  ordering on outgoing edges is just a limited version of edge labelling.} Node
labels consist of two parts: the \emph{type} and the \emph{parameters}. The type
of a node determines its out-degree.

Along with many node types defined previously
\citep{DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11}, \textsc{Crane}
uses three more:
\begin{itemize}
  \item a type for constraint removal denoted by $\CR$,
  \item a type for generalised domain recursion denoted by $\GDR$ (both with
        out-degree one),
  \item and $\star$---a placeholder type (with out-degree zero) for nodes that
        are going to be replaced.
\end{itemize}
Similarly to \citet{DBLP:conf/ijcai/BroeckTMDR11}, we write $T_p$ for an FCG
that has a node with the label $T_p$ (i.e., type $T$ and parameter(s) $p$) and
$\star$'s as all of its direct successors. See \cref{fig:examplefcg} for an
example FCG\@. Its source node has out-degree 1, label
$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$, and type $\GDR$.

% Connecting FCGs with formulas
Finally, we introduce a structure that represents a solution to a (W)FOMC
problem while it is still being built. A \emph{chip} is a pair $(G, L)$, where
$G$ is an FCG, and $L$ is a list of formulas, s.t. $|L|$ is equal to the number
of $\star$'s in $G$. $L$ contains formulas that still need to be compiled. Once
a formula is compiled, it replaces one of the $\star$'s in $G$ according to a
set order. We say that an FCG is \emph{complete} (i.e., it represents a
\emph{complete solution}) if it has no $\star$'s. Similarly, a chip is complete
if its FCG is complete (or, equivalently, the list of formulas is empty).

@@

A \emph{(compilation) rule} takes a formula and returns a set of chips. The
cardinality of this set is the number of different ways in which the rule can be
applied to the input formula. While \textsc{ForcLift}
\citep{DBLP:conf/ijcai/BroeckTMDR11} heuristically chooses one of them, in an
attempt to not miss a solution, \textsc{Crane} returns them all. In particular,
if a rule returns an empty set, then that rule does not apply to the formula.

@@

% Description of GDR, in contrast to DR
In contrast, GDR has only one precondition: for GDR to be applicable on domain
$\Omega \in \mathcal{D}$, there must be at least one variable with domain
$\Omega$ that is featured in a literal (and not just in constraints). Without
such variables, GDR would have no effect on the formula. GDR is also simpler in
that the expanded formula is left as-is to be handled by other compilation
rules. Typically, after a few more rules are applied, a combination of $\CR$ and
$\Reff$ nodes introduces a cycle-inducing edge back to the $\GDR$ node, thus
completing the definition of a recursive function. The GDR compilation rule is
summarised as \cref{alg:domainrecursion} and explained in more detail using the
example below.

@@

\begin{example}
  Let $\phi = \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (X, x), (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
          \end{multlined}\\
    c_2 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \\
            \{\, (X, x), (Z, X), (Z, x) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Gamma \,\}).
          \end{multlined}
  \end{align*}
  Domain $\Gamma$ and its element $x \in \Gamma$ satisfy the preconditions for
  CR\@. The rule introduces a new domain $\Gamma'$ and transforms $\phi$ to
  $\phi' = (c_1', c_2')$, where
  \begin{align*}
    c_1' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
           \end{multlined} \\
    c_2' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (Z, X) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Gamma' \,\}).
           \end{multlined}
  \end{align*}
\end{example}

@@

\paragraph{Notation.}
Let $\Doms$ be a function that maps any clause or formula to the set of domains
used within, and let $\dom(f)$ denote the domain of function $f$.

\paragraph{Hashing.}
We use (integer-valued) hash functions to efficiently discard pairs of formulas
that are too different for recursion to be established. The hash code of a
clause $c = (L, C, \delta_{c})$ (denoted by $\# c$) combines the hash codes of
the sets of constants and predicates in $c$, the numbers of positive and
negative literals, the number of inequality constraints $|C|$, and the number of
variables $|\Vars(c)|$. The hash code of a formula $\phi$ combines the hash
codes of all its clauses and is denoted by $\#\phi$.

\paragraph{Caching.}
\textsc{ForcLift} \citep{DBLP:conf/ijcai/BroeckTMDR11} uses a cache to check if
a formula is identical to one of the formulas that have already been fully
compiled. To facilitate recursion, we extend the caching scheme to include
formulas that have been encountered but not fully compiled yet. Formally, we
define a \emph{cache} to be a map from integers (e.g., hash codes) to sets of
pairs of the form $(\phi, v)$, where $\phi$ is a formula, and $v$ is an FCG
node.

@@

When \textsc{ForcLift} \citep{DBLP:conf/ijcai/BroeckTMDR11} compiles a WFOMC
instance into a circuit, each node type encodes an arithmetic operation on its
inputs and parameters. These operations are then immediately performed while
traversing the circuit and using domain sizes and weights as the initial inputs.
With \textsc{Crane}, the interpretation of an FCG is a collection of functions.
Each function has (some) domain sizes as parameters and may contain recursive
calls to other functions, including itself. While there may be any number of
subsidiary functions, there is always one main function that can be called with
the sizes of the domains of the input formula as arguments. Henceforth, this
function is always called $f$, and it is defined by the source node.

@@

\begin{example}\label{example:interpretation}
  Consider the FCG from \cref{fig:examplefcg}. The input formula (i.e., the
  formula from \cref{example:first}) has two domains: $\Gamma$ and $\Delta$.
  Thus, the interpretation of the FCG is a function
  $f\colon \mathbb{N}_{0} \times \mathbb{N}_{0} \to \mathbb{R}_{\ge 0}$. Let
  $m \coloneqq |\Gamma|$, and $n \coloneqq |\Delta|$. The node labelled
  $\bigvee_{\Delta^{\top} \subseteq \Delta}$ tells us that
  $f(m, n) = \sum_{l = 0}^{n} \binom{n}{l} \square$, where $\square$ is the
  interpretation of the remaining subgraph, and $l$ iterates over all possible
  sizes of $\Delta^{\top}$. It also creates two subdomains
  $\Delta^{\top}, \Delta^{\bot} \subseteq b$ that partition $\Delta$, i.e., as
  the size of $\Delta^{\top}$ increases, the size of $\Delta^{\bot}$
  correspondingly decreases. Nodes labelled $\land$ correspond to
  multiplication. Therefore,
  $f(m, n) = \sum_{l = 0}^{n} \binom{n}{l} \diamondsuit \times \heartsuit$,
  where $\diamondsuit$ is the interpretation of the contradiction (i.e., $\bot$)
  node, and $\heartsuit$ is the interpretation of the $\Reff$ node.

@@

We compare \textsc{Crane} and \textsc{ForcLift}
\citep{DBLP:conf/ijcai/BroeckTMDR11} on their ability to count various kinds of
functions. (Other WFOMC algorithms such as \textsc{L2C}
\citep{DBLP:conf/kr/KazemiP16} and probabilistic theorem proving
\citep{DBLP:journals/cacm/GogateD16} are unable to solve any of the instances
that \textsc{ForcLift} fails on.) We begin by describing how such functions can
be expressed in FOL\@. \textsc{ForcLift} then translates these sentences in FOL
to formulas as defined in \cref{def:formula}.
